Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-11-24 concurrency/CPP 2024/paper.tex
WIP

@@

\lstdefinestyle{myStyle}{
	backgroundcolor=\color{white},   
	commentstyle=\color{codegreen},
	keywordstyle=\bfseries\color{green!40!black},
	commentstyle=\itshape\color{purple!40!black},
	stringstyle=\color{codepurple},
	basicstyle=\ttfamily\footnotesize,
	breakatwhitespace=false,         
	breaklines=true,                 
	keepspaces=true,                 
	numbers=left,       
	numbersep=5pt,                  
	showspaces=false,                
	showstringspaces=false,
	showtabs=false,                  
	tabsize=2,
}

@@

As an example, a sequential stack \lstinline{push} operation could conventionally be specified as 
$$
 {\color{blue} \{\mathsf{stack}\ s\ p\} } \ \texttt{push}(s, v)\  {\color{blue} \{\mathsf{stack}\ (v :: s)\ p\}}$$  Its concurrent counterpart could be specified as 
\begin{mathpar}
	{\color{blue} \fforall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\ } 
	 \vspace{-0.8em}  \\ \texttt{push}(s, v)\   \vspace{-0.8em}  \\ 
	{\color{blue} \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle }
\end{mathpar}
indicating that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution. The stack itself is a shared resource, and can only be accessed and modified atomically by threads holding the corresponding \texttt{is\_stack} assertion. The local and shared assertions are connected by an arbitrary identifier $g$, which we will generally omit when clear from context. Logical atomicity can be used to prove linearizability---if all of a data structure's operations satisfy logically atomic triples derived from the corresponding sequential operations, the data structure is linearizable~\cite{la-lin}---but can also be used to specify more complex behavior of nonlinearizable data structures~\cite{compass}.

@@

A concurrent search structure is a data structure that supports three operations: \lstinline{insert}, \lstinline{lookup}, and \lstinline{delete}. It is intended to implement a map from keys to values, and to behave correctly when accessed simultaneously by any number of threads. In separation logic, we can prove correctness of a search structure by showing that its operations satisfy the following logically atomic specifications, where $\nodeboxrep$ is a per-thread handle to the data structure and $\treerep$ is a shared assertion linking the values in memory to an abstract map $m$ from keys to values:
\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\right\rangle
	} \vspace{-0.85em} \\ \texttt{insert(r, k, v)}\ \vspace{-0.85em} \\
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\\
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\right\rangle
	} \vspace{-0.8em}
	\\ \texttt{lookup(r, k)}\  \vspace{-0.8em} \\
	{\color{blue}
		\left\langle \texttt{v}. \ \
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m) \land m(\texttt{k}) = \texttt{v}
		\right\rangle
	}
	\\
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\right\rangle
	}  \vspace{-0.8em} 
	\\ \texttt{delete(r, k)}\ \vspace{-0.8em}\\
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
		\right\rangle
	}
\end{mathpar}
In other words, each operation takes effect atomically on the abstract map $m$, reading and/or updating it as appropriate to the operation. Krishna et al. use flow interfaces~\cite{krishna2017flow} to implement the $\treerep$ assertion connecting the data structure implementation to the abstract map, but this is not fundamental to the approach: any definition of $\nodeboxrep$ and $\treerep$ that can be used to prove the specifications (and is nontrivial, i.e., holds on the initial state of the data structure) will yield a correct data structure.

@@

\begin{mathpar}
	{\color{blue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ (m) \right\rangle}\ 
	 \vspace{-0.8em} \\ \texttt{acquire}\left(n\texttt{->lock}\right)\  \vspace{-0.8em}  \\ {\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m)\right\rangle}
\\
	{\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m) \right\rangle}\ 
	 \vspace{-0.8em} \\ \texttt{release}(n\texttt{->lock})\  \vspace{-0.8em} \\ {\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ (m)\right\rangle}
\end{mathpar}

@@

%\wm{The fonts are still off here---for instance, p, n, and n' are mathematical variables, not references to the C code. Fix if we have time.}
The \lstinline{traverse} function can then be specified as follows:
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
\begin{mathpar}
{ \color{blue}
	\fforall \  m. \ \scaleleftright[1.4ex]{\langle}
	{\begin{array}{l}
			\texttt{pn} \mapsto (p, n) \ \ast \ 
			\nodeboxrep(n)  \ast \ \mathsf{R}(n) 
		\end{array}  
		\Big \vert  \treerep\ (m)
	}  
	{\rangle}
}
\vspace{-6pt}\\
\vspace{-6pt}\texttt{traverse(pn, k)} \\
{ \color{blue}
 \scaleleftright[1.4ex]{\langle}
{\begin{array}{l}
		\mathit{res}. \exists \  n', v, \mathsf{range}. \\ 
		\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \nodeboxrep\left(n'\right) \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
		\\ 
		\ \ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ 
		\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
\end{array}  
\Bigg \vert \treerep\ (m)
}  
{\rangle}
}
\end{mathpar}

@@

The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \\ \exists \ p, n, c, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \ \\  \nodeboxrep(n)  \ \ast \  &\mathsf{node\_contents}(n, c, \mathsf{range})
\end{align*}
\begin{figure*}[!ht]
	$\color{blue}
	\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ (m) \
	\right\rangle$
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  n', v.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
		\\ 
		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
	\end{array}
	\ \Bigg| \treerep\ (m) \
	\right\rangle$
	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
	\label{proof_lock_traverse}
\end{figure*}

@@

The \lstinline{insert} function uses this specification of \lstinline{traverse} to update the state of the data structure. The desired specification of \texttt{insert} is:
\begin{mathpar}
{\color{blue}
	\fforall m.\left\langle 
	\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
	\right\rangle
} \vspace{-0.85em} \\ \texttt{insert(r, k, v)}\ \vspace{-0.85em} \\
{\color{blue}
	\left\langle 
	\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
	\right\rangle
}
\end{mathpar}
where $m$ and $m[\texttt{k} \mapsto \texttt{v}]$ denote the tree's abstract states before and after the function's execution, respectively. The proof of \lstinline{insert} is outlined in Appendix \ref{sec:apd_proof}, but informally it is quite simple. We begin by initializing the \lstinline{pn} struct with the root node and acquiring its lock, allowing us to satisfy the precondition of \lstinline{traverse}. If \lstinline{traverse} returns 0, we have found a node with key \lstinline{k}, and all we need to do is update that node's value to \lstinline{v} (line 6 in Figure \ref{insert_lock}). Otherwise, \texttt{traverse} has reached an empty node with \lstinline{k} in its range. We then call the data-structure-specific \texttt{insertOp} function, which inserts the new key-value pair at the empty node. In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}. We do this by demonstrating that in both cases, the update to the concrete data structure corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the data structure.
Finally, we release the lock acquired by \lstinline{traverse} and deallocate the \lstinline{pn} structure.

@@

The \lstinline{inRange} function raises an interesting question about the template approach: is \lstinline{inRange} part of the give-up template, or the underlying data structure? Some data structures may already track the range of keys expected in the current node, and so might define \lstinline{inRange} even in sequential implementations. However, in most sequential settings ranges can be computed from the data structure (e.g., in a binary search tree, the left subtree of a node with key \lstinline{k} holds keys less than \lstinline{k}), and there is no reason to explicitly store a node's range in the node itself. The give-up template of Krishna et al.~implicitly assumes that the underlying data structure supports \lstinline{inRange}; we prefer to consider \lstinline{inRange} part of the template, so that the template can be applied to data structures without modifying them. Accordingly, in this template we add two new fields, \lstinline{min} and \lstinline{max}, to the type \lstinline{node_t}:
\begin{lstlisting}[style=myStyle,  numbers=none, xleftmargin=0.5em]
	typedef struct node_t {
		node *t; 
		lock_t *lock; 
		int min, max;
	} node_t;
\end{lstlisting}
These fields store lower and upper bounds on the keys reachable from the current node. %In the context of binary search trees, each node is labeled with a range that is associated with a lower and upper bound on the keys. This begins with $(\texttt{min}, \texttt{max}) = (-\infty, +\infty)$ at the root node and is propagated to the empty leaf nodes. If the root node (which also serves as the parent of a node) has a key of $42$, the left leaf node's range would be $(\texttt{min}, \texttt{max}) = (-\infty, 42)$, and the right leaf node's range would be $(\texttt{min}, \texttt{max}) = (42, +\infty)$.
We can then define \texttt{inRange} as a helper function in the template, rather than requiring data structures to provide it. %\wm{This is the first but not the last implicit assumption in the templates; can we say something more general?}
Our specification of \lstinline{inRange} is:
\begin{mathpar}
	% inRange for give-up
	{\color{blue}
		\left\{ 
		\begin{array}{c}
			\texttt{pn->n->min}\mapsto n_1 \ast \texttt{pn->n->max}\mapsto n_2
		\end{array}
		\right\}
	} \vspace{-0.85em} 
	\\ \texttt{inRange(pn, k)} 
	\vspace{-0.85em}  \\
	{\color{blue}
		\left\{\mathit{res.} 
		\begin{array}{c}
			\texttt{pn->n->min}\mapsto n_1 \ast \texttt{pn->n->max}\mapsto n_2 \ \ast  
			\\ \mathsf{if}\ \mathit{res}\ \mathsf{then}\ (n_1 < \texttt{k} < n_2)\ \mathsf{else}\ (\texttt{k} \leq n_1 \lor \texttt{k} \geq n_2)
		\end{array}
		\right\}
	}
\end{mathpar}
It simply computes whether the input key \lstinline{k} is in the range $(n_1, n_2)$ associated with the node \texttt{n}, and returns 0 or 1 accordingly.

\subsubsection{Verifying the Give-Up Template}
\label{traverse_proof_giveup}
The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
\begin{mathpar}
	{ \color{blue}
		\fforall \  m. \ \scaleleftright[1.4ex]{\langle}
		{\begin{array}{l}
				\texttt{pn} \mapsto (p, n) \ast  
				\nodeboxrep(n)   
			\end{array}  
			\Big \vert  \treerep\ (m)
		}  
		{\rangle}
	}
	\vspace{-6pt}\\
	\vspace{-6pt}\texttt{traverse(pn, k)} \\
	{ \color{blue}
		\scaleleftright[1.4ex]{\langle}
		{\begin{array}{l}
				\mathit{res}. \exists \  n', v, \mathsf{range}. \\ 
				\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \nodeboxrep\left(n'\right) \ast \texttt{k} \in \mathsf{range} \ \ast 
				\\ 
				\ \ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
				\\ 
				\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
			\end{array}  
			\Bigg \vert \treerep\ (m)
		}  
		{\rangle}
	}
\end{mathpar}

@@

\subsection{Instantiating \lstinline{findNext} and \lstinline{insertOp}}
We instantiate the templates with the binary search tree by defining and verifying the helper functions \texttt{findNext} (used in \lstinline{traverse}) and \texttt{insertOp} (used in \texttt{insert}). As expected, their implementations are strictly sequential, their specifications are ordinary non-atomic Hoare triples (see Figures \ref{fig:findNext_lock} and \ref{fig:insertOp}), and their proofs do not require any concurrency reasoning.
\begin{figure*}[!ht]
	\centering
	\begin{mathpar}
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\texttt{pn->n->t}\mapsto (k', v', l, r)
			\end{array}
			\right\}
		} \vspace{-0.85em} 
		\\ 
		\texttt{findNext(pn, k)} 
		\vspace{-0.85em} 
		\\
		{\color{blue}
			\left\{\mathit{res}. \ \exists \  n'.
			\begin{array}{c}
				\texttt{pn->n->t}\mapsto (k', v', l, r)  \ \ast \\\  
				\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (l = n' \land \texttt{k} < k') \lor (r = n' \land \texttt{k} > k')\ \mathsf{else}\ (n = n' \land \texttt{k} = k')
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specification of \texttt{findNext} for the binary search tree (all templates)}
	\label{fig:findNext_lock}
\end{figure*}

>>>>

\definecolor{specblue}{rgb}{0.0,0.3,0.55}

@@

\lstdefinestyle{myStyle}{
	backgroundcolor=\color{white}, 
	keywordstyle=\bfseries\color{green!40!black},
	commentstyle=\itshape\color{violet!40!black},
	stringstyle=\color{codepurple},
	basicstyle=\ttfamily\footnotesize,
	breakatwhitespace=false,         
	breaklines=true,                 
	keepspaces=true,                 
	numbers=left,       
	numbersep=5pt,                  
	showspaces=false,                
	showstringspaces=false,
	showtabs=false,                  
	tabsize=2,
}

@@

As an example, a sequential stack \lstinline{push} operation could conventionally be specified as 
$$
 {\color{specblue} \{\mathsf{stack}\ s\ p\} } \ \texttt{push}(s, v)\  {\color{specblue} \{\mathsf{stack}\ (v :: s)\ p\}}$$  Its concurrent counterpart could be specified as 
\begin{mathpar}
	{\color{specblue} \fforall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\ } 
	 \vspace{-0.8em}  \\ \texttt{push}(s, v)\   \vspace{-0.8em}  \\ 
	{\color{specblue} \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle }
\end{mathpar}
indicating that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution. The stack itself is a shared resource, and can only be accessed and modified atomically by threads holding the corresponding \texttt{is\_stack} assertion. The local and shared assertions are connected by an arbitrary identifier $g$, which we will generally omit when clear from context. Logical atomicity can be used to prove linearizability---if all of a data structure's operations satisfy logically atomic triples derived from the corresponding sequential operations, the data structure is linearizable~\cite{la-lin}---but can also be used to specify more complex behavior of nonlinearizable data structures~\cite{compass}.

@@

A concurrent search structure is a data structure that supports three operations: \lstinline{insert}, \lstinline{lookup}, and \lstinline{delete}. It is intended to implement a map from keys to values, and to behave correctly when accessed simultaneously by any number of threads. In separation logic, we can prove correctness of a search structure by showing that its operations satisfy the following logically atomic specifications, where $\nodeboxrep$ is a per-thread handle to the data structure and $\treerep$ is a shared assertion linking the values in memory to an abstract map $m$ from keys to values:
\begin{mathpar}
	{\color{specblue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\right\rangle
	} \vspace{-0.85em} \\ \texttt{insert(r, k, v)}\ \vspace{-0.85em} \\
	{\color{specblue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\\
	{\color{specblue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\right\rangle
	} \vspace{-0.8em}
	\\ \texttt{lookup(r, k)}\  \vspace{-0.8em} \\
	{\color{specblue}
		\left\langle \texttt{v}. \ \
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m) \land m(\texttt{k}) = \texttt{v}
		\right\rangle
	}
	\\
	{\color{specblue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\right\rangle
	}  \vspace{-0.8em} 
	\\ \texttt{delete(r, k)}\ \vspace{-0.8em}\\
	{\color{specblue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
		\right\rangle
	}
\end{mathpar}
In other words, each operation takes effect atomically on the abstract map $m$, reading and/or updating it as appropriate to the operation. Krishna et al. use flow interfaces~\cite{krishna2017flow} to implement the $\treerep$ assertion connecting the data structure implementation to the abstract map, but this is not fundamental to the approach: any definition of $\nodeboxrep$ and $\treerep$ that can be used to prove the specifications (and is nontrivial, i.e., holds on the initial state of the data structure) will yield a correct data structure.

@@

\begin{mathpar}
	{\color{specblue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ (m) \right\rangle}\ 
	 \vspace{-0.8em} \\ \texttt{acquire}\left(n\texttt{->lock}\right)\  \vspace{-0.8em}  \\ {\color{specblue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m)\right\rangle}
\\
	{\color{specblue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m) \right\rangle}\ 
	 \vspace{-0.8em} \\ \texttt{release}(n\texttt{->lock})\  \vspace{-0.8em} \\ {\color{specblue}\left\langle \nodeboxrep(n) \ |\ \treerep\ (m)\right\rangle}
\end{mathpar}

@@

%\wm{The fonts are still off here---for instance, p, n, and n' are mathematical variables, not references to the C code. Fix if we have time.}
The \lstinline{traverse} function can then be specified as follows:
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
\begin{mathpar}
{ \color{specblue}
	\fforall \  m. \ \scaleleftright[1.4ex]{\langle}
	{\begin{array}{l}
			\texttt{pn} \mapsto (p, n) \ \ast \ 
			\nodeboxrep(n)  \ast \ \mathsf{R}(n) 
		\end{array}  
		\Big \vert  \treerep\ (m)
	}  
	{\rangle}
}
\vspace{-6pt}\\
\vspace{-6pt}\texttt{traverse(pn, k)} \\
{ \color{specblue}
 \scaleleftright[1.4ex]{\langle}
{\begin{array}{l}
		\mathit{res}. \exists \  n', v, \mathsf{range}. \\ 
		\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \nodeboxrep\left(n'\right) \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
		\\ 
		\ \ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ 
		\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
\end{array}  
\Bigg \vert \treerep\ (m)
}  
{\rangle}
}
\end{mathpar}

@@

The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \\ \exists \ p, n, c, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \ \\  \nodeboxrep(n)  \ \ast \  &\mathsf{node\_contents}(n, c, \mathsf{range})
\end{align*}
\begin{figure*}[!ht]
	$\color{specblue}
	\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ (m) \
	\right\rangle$
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
	$\color{specblue}
	\left\langle \mathit{res}. \ \exists \  n', v.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
		\\ 
		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
	\end{array}
	\ \Bigg| \treerep\ (m) \
	\right\rangle$
	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
	\label{proof_lock_traverse}
\end{figure*}

@@

The \lstinline{insert} function uses this specification of \lstinline{traverse} to update the state of the data structure. The desired specification of \texttt{insert} is:
\begin{mathpar}
{\color{specblue}
	\fforall m.\left\langle 
	\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
	\right\rangle
} \vspace{-0.85em} \\ \texttt{insert(r, k, v)}\ \vspace{-0.85em} \\
{\color{specblue}
	\left\langle 
	\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
	\right\rangle
}
\end{mathpar}
where $m$ and $m[\texttt{k} \mapsto \texttt{v}]$ denote the tree's abstract states before and after the function's execution, respectively. The proof of \lstinline{insert} is outlined in Appendix \ref{sec:apd_proof}, but informally it is quite simple. We begin by initializing the \lstinline{pn} struct with the root node and acquiring its lock, allowing us to satisfy the precondition of \lstinline{traverse}. If \lstinline{traverse} returns 0, we have found a node with key \lstinline{k}, and all we need to do is update that node's value to \lstinline{v} (line 6 in Figure \ref{insert_lock}). Otherwise, \texttt{traverse} has reached an empty node with \lstinline{k} in its range. We then call the data-structure-specific \texttt{insertOp} function, which inserts the new key-value pair at the empty node. In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}. We do this by demonstrating that in both cases, the update to the concrete data structure corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the data structure.
Finally, we release the lock acquired by \lstinline{traverse} and deallocate the \lstinline{pn} structure.

@@

The \lstinline{inRange} function raises an interesting question about the template approach: is \lstinline{inRange} part of the give-up template, or the underlying data structure? Some data structures may already track the range of keys expected in the current node, and so might define \lstinline{inRange} even in sequential implementations. However, in most sequential settings ranges can be computed from the data structure (e.g., in a binary search tree, the left subtree of a node with key \lstinline{k} holds keys less than \lstinline{k}), and there is no reason to explicitly store a node's range in the node itself. The give-up template of Krishna et al.~implicitly assumes that the underlying data structure supports \lstinline{inRange}; we prefer to consider \lstinline{inRange} part of the template, so that the template can be applied to data structures without modifying them. Accordingly, in this template we add two new fields, \lstinline{min} and \lstinline{max}, to the type \lstinline{node_t}:
\begin{lstlisting}[style=myStyle,  numbers=none, xleftmargin=0.5em]
	typedef struct node_t {
		node *t; 
		lock_t *lock; 
		int min, max;
	} node_t;
\end{lstlisting}
These fields store lower and upper bounds on the keys reachable from the current node. %In the context of binary search trees, each node is labeled with a range that is associated with a lower and upper bound on the keys. This begins with $(\texttt{min}, \texttt{max}) = (-\infty, +\infty)$ at the root node and is propagated to the empty leaf nodes. If the root node (which also serves as the parent of a node) has a key of $42$, the left leaf node's range would be $(\texttt{min}, \texttt{max}) = (-\infty, 42)$, and the right leaf node's range would be $(\texttt{min}, \texttt{max}) = (42, +\infty)$.
We can then define \texttt{inRange} as a helper function in the template, rather than requiring data structures to provide it. %\wm{This is the first but not the last implicit assumption in the templates; can we say something more general?}
Our specification of \lstinline{inRange} is:
\begin{mathpar}
	% inRange for give-up
	{\color{specblue}
		\left\{ 
		\begin{array}{c}
			\texttt{pn->n->min}\mapsto n_1 \ast \texttt{pn->n->max}\mapsto n_2
		\end{array}
		\right\}
	} \vspace{-0.85em} 
	\\ \texttt{inRange(pn, k)} 
	\vspace{-0.85em}  \\
	{\color{specblue}
		\left\{\mathit{res.} 
		\begin{array}{c}
			\texttt{pn->n->min}\mapsto n_1 \ast \texttt{pn->n->max}\mapsto n_2 \ \ast  
			\\ \mathsf{if}\ \mathit{res}\ \mathsf{then}\ (n_1 < \texttt{k} < n_2)\ \mathsf{else}\ (\texttt{k} \leq n_1 \lor \texttt{k} \geq n_2)
		\end{array}
		\right\}
	}
\end{mathpar}
It simply computes whether the input key \lstinline{k} is in the range $(n_1, n_2)$ associated with the node \texttt{n}, and returns 0 or 1 accordingly.

\subsubsection{Verifying the Give-Up Template}
\label{traverse_proof_giveup}
The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
\begin{mathpar}
	{ \color{specblue}
		\fforall \  m. \ \scaleleftright[1.4ex]{\langle}
		{\begin{array}{l}
				\texttt{pn} \mapsto (p, n) \ast  
				\nodeboxrep(n)   
			\end{array}  
			\Big \vert  \treerep\ (m)
		}  
		{\rangle}
	}
	\vspace{-6pt}\\
	\vspace{-6pt}\texttt{traverse(pn, k)} \\
	{ \color{specblue}
		\scaleleftright[1.4ex]{\langle}
		{\begin{array}{l}
				\mathit{res}. \exists \  n', v, \mathsf{range}. \\ 
				\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \nodeboxrep\left(n'\right) \ast \texttt{k} \in \mathsf{range} \ \ast 
				\\ 
				\ \ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
				\\ 
				\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
			\end{array}  
			\Bigg \vert \treerep\ (m)
		}  
		{\rangle}
	}
\end{mathpar}

@@

\subsection{Instantiating \lstinline{findNext} and \lstinline{insertOp}}
We instantiate the templates with the binary search tree by defining and verifying the helper functions \texttt{findNext} (used in \lstinline{traverse}) and \texttt{insertOp} (used in \texttt{insert}). As expected, their implementations are strictly sequential, their specifications are ordinary non-atomic Hoare triples (see Figures \ref{fig:findNext_lock} and \ref{fig:insertOp}), and their proofs do not require any concurrency reasoning.
\begin{figure*}[!ht]
	\centering
	\begin{mathpar}
		{\color{specblue}
			\left\{ 
			\begin{array}{c}
				\texttt{pn->n->t}\mapsto (k', v', l, r)
			\end{array}
			\right\}
		} \vspace{-0.85em} 
		\\ 
		\texttt{findNext(pn, k)} 
		\vspace{-0.85em} 
		\\
		{\color{specblue}
			\left\{\mathit{res}. \ \exists \  n'.
			\begin{array}{c}
				\texttt{pn->n->t}\mapsto (k', v', l, r)  \ \ast \\\  
				\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (l = n' \land \texttt{k} < k') \lor (r = n' \land \texttt{k} > k')\ \mathsf{else}\ (n = n' \land \texttt{k} = k')
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specification of \texttt{findNext} for the binary search tree (all templates)}
	\label{fig:findNext_lock}
\end{figure*}

\begin{figure*}[h]
	\begin{subfigure}{\textwidth}
		\centering
		\begin{mathpar}
			{\color{specblue}
				\left\{ 
				\begin{array}{c}
					\ \texttt{pn->n} \mapsto \texttt{NULL}  
				\end{array}
				\right\}
			} \vspace{-0.85em} 
			\\ 
			\texttt{insertOp(pn, k, v)} 
			\vspace{-0.85em} 
			\\
			{\color{specblue}
				\left\{\exists \  l\ r\ \mathit{lk}_1\ \mathit{lk}_2.
				\begin{array}{l}
					\ \texttt{pn->n->t} \mapsto (\texttt{k}, \texttt{v}, l, r) \ \ast \ \\ 
					\ l \mapsto (\texttt{NULL}, \mathit{lk}_1)  \ast  r \mapsto (\texttt{NULL}, \mathit{lk}_2)
				\end{array}
				\right\}
			}
		\end{mathpar}
		\caption{Specification of \texttt{insertOp} for the lock-coupling template}
		\label{fig:insertOp_lock}
	\end{subfigure}
	%insertOp for give-up of give-up
	\begin{subfigure}{\textwidth}
		\begin{mathpar}
			{\color{specblue}
				\left\{ 
				\begin{array}{c}
					\texttt{pn->n}\mapsto \texttt{NULL}  \ast  \texttt{pn->n->min}\mapsto n_1  \ast  \texttt{pn->n->max}\mapsto n_2
				\end{array}
				\right\}
			} \vspace{-0.85em} 
			\\
			\texttt{insertOp(pn, k, v)}
			\vspace{-0.85em} 
			\\
			{\color{specblue}
				\left\{\exists \ l\ r\ \mathit{lk}_1\ \mathit{lk}_2. 
				\begin{array}{c}
					\texttt{pn->n->t}\mapsto (\texttt{k}, \texttt{v}, l, r)  \ast  \texttt{pn->n->min}\mapsto n_1 \ast  \texttt{pn->n->max}\mapsto n_2  \ \ast \ 
					\\ l \mapsto (\texttt{NULL}, \mathit{lk}_1, (n_1, \texttt{k})) \ast  r \mapsto (\texttt{NULL}, \mathit{lk}_2, (\texttt{k}, n_2))
				\end{array}
				\right\}
			}
		\end{mathpar}
		\caption{Specification of \texttt{insertOp} for the give-up template}
		\label{fig:insertOp_giveup}
	\end{subfigure}
	\caption{Specification of \texttt{insertOp} for the lock-coupling and give-up templates}
	\label{fig:insertOp}
\end{figure*}
